What paradigms can webcam eye-tracking be used for? Attempted replications of five ``classic'' cognitive science experiments

### 2024-10-24 group-b/E2-analysis-replication.Rmd
A bunch of edits through Experiment 2

The proportions of looks across quadrants in the free-viewing condition were analyzed using a linear mixed-effects model with quadrant as the predictor (critical as the reference level). The model included random intercepts and slopes for participants^[ `lme4` syntax: `lmer(proportion ~ quadrant + (1+quadrant|subject_id))`. Among other limitations, this approach violates the independence assumptions of the linear model because looks to the four locations are not independent. This analysis was chosen because it is analogous to the ANOVA analysis conducted in the original paper.]. Proportions of looks were significantly higher for the critical quadrant compared to the other three (first: _b_ =  `r E2_gaze_model_q1 %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_gaze_model_q1 %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001, second: _b_ =  `r E2_gaze_model_q2 %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_gaze_model_q2 %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001, third: _b_ =  `r E2_gaze_model_q3 %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_gaze_model_q3 %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001).

@@

### Response Time and Accuracy
Participants' response times and accuracies on memory questions are summarized in Figure \@ref(fig:E2-rt-acc-fig). 
Both dependent variables were analyzed with linear mixed-effects model with relation type (interobject = -0.5, intraobject=0.5) and viewing_condition (fixed = -0.5, free=0.5) and their interaction as the predictors. 
The model included random intercepts for participants^[ `lme4` syntax: `lmer(DV ~ relation_type*viewing_condition + (1|subject_id))`]. 
Accuracy did not differ significantly between interobject and intraobject questions (_b_ =  `r E2_acc_model_rel %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_acc_model_rel %>% pull(std.error) %>% round(digits = 2)`, _p_=`r E2_acc_model_rel %>% pull(p.value) %>% round(digits = 2)`). Participants were less accurate in the free viewing condition than the fixed condition (_b_ =  `r E2_acc_model_cond %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_acc_model_cond %>% pull(std.error) %>% round(digits = 2)`, _p_=`r E2_acc_model_cond %>% pull(p.value) %>% round(digits = 2)`). 
Response times were slower for interobject (e.g., "The train is to the right of the taxi.") than intraobject (e.g., "The train is facing right.") questions (_b_ =  `r E2_RT_model_rel %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_RT_model_rel %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001). Response times were slower in the free viewing condition than the fixed condition (_b_ =  `r E2_RT_model_cond %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_RT_model_cond %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001). 
The interaction was not a significant predictor for response times or accuracy.
These behavioral results are inconsistent with the original findings.

@@

One possibility is that the in-lab participants in @johanssonLookHereEye2014 were much more compliant with the instruction to keep their gaze on central fixation (though these data are not reported in the original paper). When  analyzing results from the subset of participants (N = `r length(successful_fixed)`) who were most compliant during the fixed-viewing block (at least 25% of their looks fell within 20% of the center of the display), the viewing condition effects and the interactions were not signficant. Given the smaller sample size we do not interpret these results further.

>>>>

The proportions of looks across quadrants in the free-viewing condition were analyzed using a linear mixed-effects model with quadrant as the predictor (critical as the reference level). The model included random intercepts and slopes for participants.^[ `lme4` syntax: `lmer(proportion ~ quadrant + (1+quadrant|subject_id))`. Among other limitations, this approach violates the independence assumptions of the linear model because looks to the four locations are not independent. This analysis was chosen because it is analogous to the ANOVA analysis conducted in the original paper.] Proportions of looks were significantly higher for the critical quadrant compared to the other three (first: _b_ =  `r E2_gaze_model_q1 %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_gaze_model_q1 %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001, second: _b_ =  `r E2_gaze_model_q2 %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_gaze_model_q2 %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001, third: _b_ =  `r E2_gaze_model_q3 %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_gaze_model_q3 %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001).

@@

### Response Time and Accuracy
Participants' response times and accuracies on memory questions are summarized in Figure \@ref(fig:E2-rt-acc-fig). Both dependent variables were analyzed with linear mixed-effects model with relation type (interobject = -0.5, intraobject=0.5) and viewing_condition (fixed = -0.5, free=0.5) and their interaction as the predictors and random intercepts for participants.^[ `lme4` syntax: `lmer(DV ~ relation_type*viewing_condition + (1|subject_id))`]

The original study reported RTs for interobject statements were longer in the fixed condition relative to the free-viewing condition, while there was no difference for intraobject statements, resulting in a significant interaction. In contrast, we found overall slower RTs for the free viewing condition (_b_ =  `r E2_RT_model_cond %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_RT_model_cond %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001) and for interobject questions (_b_ =  `r E2_RT_model_rel %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_RT_model_rel %>% pull(std.error) %>% round(digits = 2)`, _p_<0.001) but no interaction. We also failed to replicate a significant main effect of question type on accuracy (_b_ =  `r E2_acc_model_rel %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_acc_model_rel %>% pull(std.error) %>% round(digits = 2)`, _p_=`r E2_acc_model_rel %>% pull(p.value) %>% round(digits = 2)`), while we found an unexpected effect of viewing condition on accuracy, with higher accuracy for the fixed condition (_b_ =  `r E2_acc_model_cond %>% pull(estimate) %>% round(digits = 2)`, _SE_ =  `r E2_acc_model_cond %>% pull(std.error) %>% round(digits = 2)`, _p_=`r E2_acc_model_cond %>% pull(p.value) %>% round(digits = 2)`).

@@

One possibility is that the in-lab participants in @johanssonLookHereEye2014 were much more compliant with the instruction to keep their gaze on central fixation (though these data are not reported in the original paper). However, restricting analyses to the subset of participants (N = `r length(successful_fixed)`) who were most compliant during the fixed-viewing block (at least 25% of their looks fell within 20% of the center of the display), the viewing condition effects and the interactions remained non-signficant. Given the smaller sample size we do not interpret these results further.

### 2024-10-24 manuscript.Rmd
A bunch of edits through Experiment 2

abstract: |
  Web-based data collection allows researchers to recruit large and diverse samples with fewer resources than lab-based studies require. Recent innovations have expanded the set of methodolgies that are possible online, but ongoing work is needed to test the suitability of web-based tools for various research paradigms. Here, we focus on webcam-based eye-tracking; we tested whether the results of five different eye-tracking experiments in the cognitive psychology literature would replicate in a webcam-based format. Specifically, we carried out five experiments by integrating two javascript-based tools: js.psych and a modified version of Webgazer.js. In order to represent a wide range of applications of eye-tracking to cognitive psychology, we chose two psycholinguistic experiments, two memory experiments, and a decision-making experiment. These studies also varied in the type of eye-tracking display, including screens split into halves (Exps. 3 and 5) or quadrants (Exps. 2 and 4), or composed scenes with regions of interest that varied in size (Exp. 1). Outcomes were mixed. The least successful replication attempt was Exp. 1; we did not obtain a condition effect in our remote sample (1a), nor in an in-lab follow-up (1b). However, the other four experiments were more successful, replicating a blank-screen effect (Exp. 2), a novelty preference (Exp. 3), a verb bias effect (Exp. 4), and a gaze-bias effect in decision-making (Exp. 5). These results suggest that webcam-based eye tracking can be used to detect a variety of cognitive phenomena, including those with sensitive time, although paradigms that require high spatial resolution should be adapted to coarser quadrant or split-half displays.

@@

A robust solution to all these problems is online experiments, particularly with volunteer citizen scientists as participants [@li2024developmental; @hartshorne2019thousand; @reinecke2015labinthewild; @gosling2010wired]. Historically, this option has not been available for eye-tracking. This began to shift with the widespread incorporation of cameras into computers; researchers have long used frame-by-frame analysis of video for low-resolution eyetracking [@snedeker2003using]. Unfortunately, this can be extremely time-intensive, making large-sample studies unrealistic. In recent years, image analysis has improved to the point where this work can be automated with reasonable accuracy [@papoutsaki2016webgazer; @zheng2018rapid; @burton2014comparison; @skovsgaard2011evaluation]. However, webcam-based eyetracking only started to be used regularly in research with the advent of `Webgazer.js` [@papoutsaki2016webgazer], a webcam-based Javascript plug-in that works in the browser and which can be integrated with any Javascript web interface, including `jsPsych` [@deleeuwJsPsychJavaScriptLibrary2015], `Gorilla` [@anwyl2020gorilla], or `lab.js` [@henninger2021lab].

Given the potential game-changing nature of webcam-based eye-tracking, a number of research groups have investigated how well it works. There are two potential limitations to webcam-based eye-tracking. First, the spatial and temporal resolution is less than what is achievable with an infrared system. Second, testing subjects over the Internet involves less control: subjects may be unable or unwilling to calibrate equipment, adjust lighting, etc., to the same level of precision typical of in-lab studies. Nearly all this work has used `Webgazer.js` --- most in combination with jsPsych, but some with Gorilla or hand-built integrations.

@@

Second, prior work has focused on studies with relatively coarse-grained regions of interest [but see @semmelmann2018online], dividing the screen either in half or in quadrants. This is particularly salient with [@prystauka2023online], which simplified [@altmannIncrementalInterpretationVerbs1999]'s design so that regions of interest are quadrants rather than the finer-grained ROIs used in the original. Certainly, webcam eyetracking will not be as spatially fine-grained as an infrared eyetracker, but we do not yet have a good sense of the limits.

@@

In order to validate the online eye-tracking methodology, with the particular configuration known to have the greatest temporal precision, `jsPsych` and a modification of `Webgazer`, we set out to reproduce five previously published studies representing a variety of questions, topics, and paradigms. The goal was to examine the strengths and weaknesses of webcam eye-tracking for common paradigms in cognitive science, across a broad range of research areas.

@@

Studies with large effect sizes and which are known to replicate are ideal targets for further replication; otherwise, it can be difficult to distinguish a failure of the method from a failure of the original study to replicate. In practice, replications (successful or otherwise) have only been reported for a small number of studies, so we ultimately included some studies with unknown replicability. We addressed this in several ways. First, replicating five very different studies from different research traditions decreases our reliance on any one study. Second, we include several ''sanity check'' analyses, such as the correlation between calibration accuracy and effect size. If the effect is real but there is noise from low-accuracy eye-tracking, this correlation should be substantial. Third, for two of the studies, we had comparison data collected in-lab either using `jsPsych` or a more traditional eyetracker technology, allowing us to directly assess the impact of differences in subject population and equipment/setting.

@@

We chose five high-impact eye-tracking studies involving adult subjects [for an investigation of WebGazer's validity for developmental research, see @steffan2024validation for a comparison of remote WebGazer and in-lab anticipatory looking effects in 18-27-month-old participants]. Our goal was to include experiments from a range of topic areas (e.g., memory, decision making, psycholinguistics) and paradigms (two halves of the screen, visual world paradigm with four quadrants, visual world paradigm with "naturalistic" scenes). As noted above, we had a preference for well-established findings that are known to replicate, though for sake of diversity this was not always possible. Table \@ref(tab:studies-table) provides an overview of the five studies we selected.

@@

We used `r cite_r("references.bib")` for all our analyses.

@@

### Participants
Sixty participants were paid \$2.60 for their participation. Our sample size
of participants was determined by the total run time of our experiment,
\~10 minutes, and the allotted funding from the Vassar College Cognitive
Science Department. From this information, we calculated a reasonable
number of participants we could afford to compensate on Prolific. Note
that the sample size of the original study was 24. For unknown reasons,
2 of the subjects' results were not recorded, so in the analysis, we
worked with data collected from 58 participants.

@@

## Discussion
Across three different tests of the hypothesis that listeners will use verb semantics to anticipate the upcoming referent, we found results that were numerically consistent with the hypothesis but not statistically significant. A comparison to published data [@james2023language] demonstrates that looks to the objects in the scene (relative to background or off-screen looks) were depressed across conditions and objects. After eliminating participants with validation accuracy under 50% and/or 10% or fewer fixations to any ROIs, we were left with only 22 of the original 60 participants. Analyses on this subset resulted in one effect reaching statistical significance (shorter target fixation latencies in the restricting vs. non-restricting verb condition). 
Given that nearly two-thirds of participants had poor data quality, we ran a follow-up webcam study in a lab setting order to isolate the potential causes.

@@

## Discussion
Overall, the results of Experiment 1 paint a sobering picture of web-based eye-tracking. In Experiment 1a, results from the remote samplewere in the expected direction but effects were smaller and delayed relative to previous work and failed to reach statistical significance. To test whether this could be explained by the variability in experimental settings across participants, we replicated our procedure in a lab setting. Surprisingly, the results in the in-lab study were less aligned with previous work; critical effects were not significant nor were they in the expected direction numerically. This suggests the problem was not with running the study online or due to working with a more diverse population via Prolific.

Taken together, the instability of these effects might suggest that this paradigm is not well-suited for webcam-based eye-tracking. Notably, the ROIs were tightly drawn around the five to six objects in each scene (drawing larger ROIs in these scenes would have led to overlapping objects) and thus, analyses were unforgiving of inaccurate calibration. Further evidence comes from a recent `webgazer` study that successfully replicated a modified version of the Altmann and Kamide study with only four objects, each in separate quadrants, allowing for larger, more distinct ROIs [@prystauka2023online].

@@

Sixty participants were paid for their participation.
The sample size was motivated in part by budget constraints, but was
nonetheless 2.5x larger than the original sample size of 24). 
Data from one participant were not properly
recorded due to unknown technical issues, so data from 59 participants
were included in all analyses to follow.

@@

As in @johanssonLookHereEye2014 and @spiveyOculomotorMechanismsActivated2001, during memory retrieval, learners spontaneously look to blank screen locations where pictures were located
during encoding, suggesting that visuospatial information is integrated into the memory
for objects. However, we did not observe a memory benefit, in terms of speed or accuracy, of spatial reinstatement via gaze position during retrieval of the picture. We can speculate that this may be due to the fact that participants struggled to maintain their gaze fixed in the center in the fixed-viewing condition, such that the difference between the fixed- and free-viewing conditions was minimal. Crucially for the current purposes, the webcam-based eye-tracking measurements were successful in replicating the key eye-tracking results.

>>>>

abstract: |
  Web-based data collection allows researchers to recruit large and diverse samples with fewer resources than lab-based studies require. Recent innovations have expanded the set of methodolgies that are possible online, but ongoing work is needed to test the suitability of web-based tools for various research paradigms. Here, we focus on webcam-based eye-tracking; we tested whether the results of five different eye-tracking experiments in the cognitive psychology literature would replicate in a webcam-based format. Specifically, we carried out five experiments by integrating two javascript-based tools: jsPsych and a modified version of Webgazer.js. In order to represent a wide range of applications of eye-tracking to cognitive psychology, we chose two psycholinguistic experiments, two memory experiments, and a decision-making experiment. These studies also varied in the type of eye-tracking display, including screens split into halves (Exps. 3 and 5) or quadrants (Exps. 2 and 4), or composed scenes with regions of interest that varied in size (Exp. 1). Outcomes were mixed. The least successful replication attempt was Exp. 1; we did not obtain a condition effect in our remote sample (1a), nor in an in-lab follow-up (1b). However, the other four experiments were more successful, replicating a blank-screen effect (Exp. 2), a novelty preference (Exp. 3), a verb bias effect (Exp. 4), and a gaze-bias effect in decision-making (Exp. 5). These results suggest that webcam-based eye tracking can be used to detect a variety of cognitive phenomena, including those with sensitive time, although paradigms that require high spatial resolution (like Exp. 1) should be adapted to coarser quadrant or split-half displays.

@@

A robust solution to all these problems is online experiments, particularly with volunteer citizen scientists as participants [@li2024developmental; @hartshorne2019thousand; @reinecke2015labinthewild; @gosling2010wired]. Historically, this option has not been available for eye-tracking, since few potential subjects have expensive eyetrackers at home. In principal, this could be circumvented by using so-called "poor-man's" eye-tracking, in which researchers videotape the subject's face and code eye position frame-by-frame [@snedeker2003using]. Most computers now have high-quality built-in cameras, so recording video is straightforward. Unfortunately, coding images frame-by-frame is extremely time-intensive, making large-sample studies unrealistic. However, in recent years, image analysis has improved to the point where this work can be automated with reasonable accuracy [@papoutsaki2016webgazer; @zheng2018rapid; @burton2014comparison; @skovsgaard2011evaluation]. Nonetheless, webcam-based eyetracking only started to be used regularly in research with the advent of `Webgazer.js` [@papoutsaki2016webgazer], a webcam-based Javascript plug-in that works in the browser and which can be integrated with any Javascript web interface, including `jsPsych` [@deleeuwJsPsychJavaScriptLibrary2015], `Gorilla` [@anwyl2020gorilla], or `lab.js` [@henninger2021lab].

Given the potential game-changing nature of webcam-based eye-tracking, a number of research groups have investigated how well it works. Nearly all this work has used `Webgazer.js` --- most in combination with jsPsych, but some with Gorilla or hand-built integrations. There are two potential limitations to webcam-based eye-tracking. First, the spatial and temporal resolution is less than what is achievable with an infrared system. This is less a limitation of webcams than a limitation of the information about eyegaze available in an image of a face. Anecdotally, researchers who use poor-man's eyetracking find it difficult to locate the gaze to anything more fine-grained than quadrants. Second, testing subjects over the Internet involves less control: subjects may be unable or unwilling to calibrate equipment, adjust lighting, etc., to the same level of precision typical of in-lab studies.

@@

Second, prior work has focused on studies with relatively coarse-grained regions of interest [but see @semmelmann2018online], dividing the screen either in half or in quadrants. This is particularly salient with [@prystauka2023online], which simplified @altmannIncrementalInterpretationVerbs1999's design so that regions of interest are quadrants rather than the finer-grained ROIs used in the original. Certainly, webcam eyetracking will not be as spatially fine-grained as an infrared eyetracker, but we do not yet have a good sense of the limits.

@@

In order to broaden the validation of online eye-tracking methodology, we set out to reproduce five previously published studies representing a variety of questions, topics, and paradigms. The goal was to examine the strengths and weaknesses of webcam eye-tracking for common paradigms in cognitive science, across a broad range of research areas. Critically, we used the implementation known to have the greatest temporal precision to date: `jsPsych` and a modification of `Webgazer`.

@@

Studies with large effect sizes and which are known to replicate are ideal targets for further replication; otherwise, it can be difficult to distinguish a failure of the method from a failure of the original study to replicate. In practice, replications (successful or otherwise) have only been reported for a small number of studies, so we ultimately included some studies with unknown replicability. We addressed this in several ways. First, replicating five very different studies from different research traditions decreases our reliance on any one study. Second, we include several "sanity check" analyses, such as the correlation between calibration accuracy and effect size. If the effect is real but there is noise from low-accuracy eye-tracking, this correlation should be substantial. Third, for two of the studies, we had comparison data collected in-lab either using `jsPsych` or a more traditional eyetracker technology, allowing us to directly assess the impact of differences in subject population and equipment/setting.

@@

We chose five high-impact eye-tracking studies involving adult subjects [for an comparison of remote WebGazer and in-lab anticipatory looking effects with 18-27 month-old participants, see @steffan2024validation]. Our goal was to include experiments from a range of topic areas (e.g., memory, decision making, psycholinguistics) and paradigms (two halves of the screen, visual world paradigm with four quadrants, visual world paradigm with "naturalistic" scenes). As noted above, we had a preference for well-established findings that are known to replicate, though for sake of diversity this was not always possible. Table \@ref(tab:studies-table) provides an overview of the five studies we selected.

@@

We used `r cite_r("references.bib")` for our analyses.

@@

### Participants
Sixty participants (2.5 times the original sample of 24) were paid \$2.60 for their participation. For unknown reasons,
2 of the subjects' results were not recorded, so in the analysis, we
worked with data collected from 58 participants.

@@

## Discussion
Across three different tests of the hypothesis that listeners will use verb semantics to anticipate the upcoming referent, we found results that were numerically consistent with the hypothesis but not statistically significant. This is unlikely to be because the original effect is unreliable. Indeed, one of us had successfully replicated it previously using traditional methods [@james2023language].

An eyeball comparison to published data [@altmannIncrementalInterpretationVerbs1999, @james2023language] suggests that looks to the objects in the scene (relative to background or off-screen looks) were depressed across conditions and objects. After eliminating participants with validation accuracy under 50% and/or 10% or fewer fixations to any ROIs, we were left with only 22 of the original 60 participants. Either of these facts could have resulted in low power to detect effects. Given that nearly two-thirds of participants had poor data quality, we ran a follow-up webcam study in a lab setting order to test whether the failure to replicate was due to using a remote sample or to the technology itself.

@@

## Discussion
Overall, the results of Experiment 1 paint a sobering picture of web-based eye-tracking. In Experiment 1a, results from the remote sample were in the expected direction but effects were smaller and delayed relative to previous work and failed to reach statistical significance. This was not due to vagaries of remote samples or working with a more diverse subject population: results from the in-lab study were, if anything, worse. Nor does our failure to replicate seem to be driven by subjects with poor calibration; effect sizes were not correlated with calibration success, and restricting analyses to subjects with good calibration scores only helped in one of the experiments and in only one of the three critical analyses.

Taken together, the instability of these effects might suggest that this paradigm is not well-suited for webcam-based eye-tracking. Notably, the ROIs were tightly drawn around the five to six objects in each scene (drawing larger ROIs in these scenes would have led to overlapping objects) and thus, analyses were unforgiving of inaccurate calibration. Further evidence comes from a recent `webgazer` study that successfully replicated the original results using modified stimuli consisting of only four objects, each in separate quadrants, allowing for larger, more distinct ROIs [@prystauka2023online].

@@

Sixty participants were paid for their participation (once again, 2.5x larger than the original sample size of 24). 
Data from one participant were not properly
recorded due to unknown technical issues, so data from 59 participants
were included in all analyses to follow.

@@

We replicated the key eyetracking result: As in @johanssonLookHereEye2014 and @spiveyOculomotorMechanismsActivated2001, during memory retrieval, learners spontaneously look to blank screen locations where pictures were located
during encoding, suggesting that visuospatial information is integrated into the memory
for objects. Interestingly, we did not replicate the behavior result: a memory benefit, in terms of speed, of spatial reinstatement via gaze position during retrieval of the picture. For researchers interested in that effect, this is an issue to be followed up on. However, it is not strictly relevant for our evaluation of gaze tracking online.
